Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model of Patient-Provider Communication Sequences

### 2017-08-28 paper-amia-summit-2018/amia_summit_2018.tex
start editing paper
update results and renamed paper title

\documentclass{amia}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{multirow}
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

@@

\title{Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences}

\author{Alexander Kotov, PhD$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Mehedi Hasan, BS$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

@@

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
Natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
Alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
  \end{tabular}
\end{table}

@@

@@

\subsection*{\textit{Predictive performance for natural PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_norm_seq}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline

 \multirow{2}{*}{Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532\\\cline{2-5}
 & Second order & 0.6889 & 0.7889 & 0.7352\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989}\\ \cline{2-5}
 & Second order & 0.7400 & \textbf{0.8449} & 0.7822\\ \hline

  \end{tabular}
\end{table}

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models. However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736). We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure}\\ \hline

 \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604\\\cline{2-5}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736}\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353\\ \cline{2-5}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699\\ \hline
  \end{tabular}
\end{table}

>>>>

\documentclass{amia_summit_2018}
\usepackage{graphicx}
\usepackage[labelfont=bf]{caption}
\usepackage[superscript,nomove]{cite}
\usepackage{color}
\usepackage{multirow}
\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

@@

\title{Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model of Patient-Provider Communication Sequences}

\author{Mehedi Hasan, BS$^{1}$\footnote[1]{Authors provided equal contribution. \label{footnote1}}, Alexander Kotov, PhD$^{1}$\textsuperscript{\ref{footnote1}}, April Idalski Carcone, PhD$^{2}$, Ming Dong, PhD$^{1}$, Sylvie Naar, PhD$^{2}$}

@@

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 880 & 82.09\% & 5.83 \\\hline
Unsuccessful sequences & 880 & 82.09\% & 5.83 \\\hline
  \end{tabular}
\end{table}

@@

\textbf {Recurrent Neural Networks (RNNs)} are a type of Neural Network where the hidden state of one time step is computed by combining the current input with the hidden state of the previous time step....

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_norm_seq}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_under_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7013 & 0.7037 & 0.7013 & 0.7003\\ \hline
 Markov Chain $2^{nd}$ Order & 0.5905 & 0.5917 & 0.5905 & 0.5890\\ \hline
 Hidden Markov Model & 0.5439 & 0.5778 & 0.5439 & 0.4865\\ \hline
 LSTM RNN & 0.8770 & 0.8812 & 0.8770 & 0.8766\\ \hline
 LSTM RNN - TR & 0.8824 & 0.8890 & 0.8824 & 0.8817\\ \hline
 GRU RNN & 0.8844 & 0.8897 & 0.8844 & 0.8840\\ \hline
 GRU RNN - TR & \textbf{0.8858} & \textbf{0.8917} & \textbf{0.8858} & \textbf{0.8853}\\ \hline 
  \end{tabular}
\end{table}

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models. However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736). We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

 Markov Chain $1^{st}$ Order & 0.7930 & 0.8127 & 0.7930 & 0.7896\\ \hline
 Markov Chain $2^{nd}$ Order & 0.7325 & 0.7461 & 0.7325 & 0.7288\\ \hline
 Hidden Markov Model & 0.7763 & 0.8062 & 0.7763 & 0.7706\\ \hline
 LSTM RNN & 0.8703 & 0.8782 & 0.8703 & 0.8696\\ \hline
 LSTM RNN - TR & 0.8713 & 0.8789 & 0.8713 & 0.8706\\ \hline
 GRU RNN & 0.8711 & 0.8785 & 0.8711 & 0.8705\\ \hline
 GRU RNN - TR & \textbf{0.8722} & \textbf{0.8790} & \textbf{0.8722} & \textbf{0.8716}\\ \hline
  \end{tabular}
\end{table}
